---
title: Applying PCA to fictional character personalities
author: Alex Cookson
date: '2020-11-20'
slug: []
categories: []
tags: []
description: ''
topics: []
---

In this post, we're going to apply Principal Component Analysis (PCA) to a dataset of fictional character personalities.

PCA is a common technique for [dimensionality reduction](https://en.wikipedia.org/wiki/Dimensionality_reduction), which you might want to do if you are, say, trying to put together a classification model and you have a dataset with *a lot* of variables.

The dataset we're using is of crowdsourced scores of personality traits for 800 fictional characters from books/movies/TV shows like *Game of Thrones*, *Pride and Prejudice*, and *The Lion King*. The data comes from the [Open-Source Psychometrics Project](https://openpsychometrics.org/)'s online ["Which Character" personality quiz](https://openpsychometrics.org/tests/characters/). You can learn more about that [here](https://github.com/tacookson/data/tree/master/fictional-character-personalities).  
\  


## Setup

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse) # For data manipulation
library(tidymodels) # For PCA
library(tidytext) # For reorder_within()
library(extrafont) # For custom fonts on graphs
library(glue) # For creating strings from multiple pieces
library(ggtext) # For fine-grained text formatting
library(plotly) # For interactive graphs

# These are to make tables look nice on the website
# You probably won't need them
library(knitr)
library(kableExtra)

theme_set(theme_void())

personalities <- read_tsv("https://raw.githubusercontent.com/tacookson/data/master/fictional-character-personalities/personalities.txt")
```  
\  


## Data inspection

Let's look at an example character: Captain Jean-Luc Picard from *Star Trek: The Next Generation*.

```{r picard-example, fig.align = 'center', fig.width = 8}
personalities %>%
  filter(character_name == "Jean-Luc Picard") %>%
  head(8) %>%
  arrange(mean) %>%
  kable(format = "html") %>%
  kable_styling()
```

The main fields we're interested in are `spectrum_low`, `spectrum_high`, and `mean`:

- The spectrum fields tell us what trait is on each end of the spectrum being considered
- `mean` is a score (from -50 to +50), where a score closer to -50 means the character is more like the `spectrum_low` trait and a score closer to +50 means the character is more like the `spectrum_high` trait

If we look at some of Captain Picard's strongest traits -- that is, close to -50 or +50 -- we see that he is **charming** and **intellectual** (low scores corresponding to `spectrum_low`), as well as **bold** and **tasteful** (high scores corresponding to `spectrum_high`). That rings true to me. I would certainly use those words to describe our dear Captain.

Most of the other fields are to identify who the character is and what fictional world they come from. We also have the number of ratings (`ratings`) and standard deviation (`sd`) for that character-spectrum combination. And finally, we have a helper column, `is_emoji` that tells us if the traits on each end of the spectrum are emojis instead of words.

Juuuust to make sure this data checks out -- and because I'm really curious -- let's look at some more examples from *Pride and Prejudice* and *The Lion King*. We'll look each character's eight strongest traits. I've included the trait on the opposite end of the spectrum in parentheses for reference.  
\  


### *Pride and Prejudice*

```{r pride-and-prejudice-example, fig.align = 'center', fig.width = 8, fig.height = 7.75}
pride_palette <- c("#19313b", "#3b636b", "#bbb6a2",
                    "#ffc277", "#bc652f", "#472411")

pride_example <- personalities %>%
  filter(!is_emoji,
         fictional_work == "Pride and Prejudice",
         character_name %in% c("Elizabeth Bennet", "Jane Bennet", "Lydia Bennet",
                               "Mr. Darcy", "Mr. William Collins", "George Wickham")) %>%
  mutate(character_name = ifelse(character_name == "Mr. William Collins", "Mr. Collins", character_name)) %>%
  group_by(character_name) %>%
  top_n(8, abs(mean)) %>%
  ungroup() %>%
  mutate(spectrum_label = ifelse(mean > 0,
                                 glue("{spectrum_high}<br><span style='font-size:8pt'>({spectrum_low})</span>"),
                                 glue("{spectrum_low}<br><span style='font-size:8pt'>({spectrum_high})</span>")),
         mean = abs(mean),
         spectrum_label = reorder_within(spectrum_label, mean, character_name))

pride_example %>%
  ggplot() +
  geom_col(aes(mean, spectrum_label, fill = character_name), show.legend = FALSE) +
  geom_vline(xintercept = seq(0, 50, by = 10), colour = "#FFF1E6", size = 0.1) +
  scale_y_reordered() +
  scale_fill_manual(values = pride_palette) +
  expand_limits(x = c(0, 50)) +
  facet_wrap(~ character_name, scales = "free_y") +
  labs(title = "Pride and Prejudice",
       caption = "Visualization: @alexcookson | Data: Open Source Psychometrics Project") +
  theme(plot.title = element_text(family = "JaneAusten", face = "bold", size = 30,
                                  hjust = 0.5, margin = margin(t = 10, b = 20)),
        plot.background = element_rect(fill = "#FFF1E6", colour = NA),
        plot.margin = margin(10, 10, 10, 10),
        plot.caption = element_text(family = "Sylfaen", size = 8, colour = "grey50", margin = margin(t = 10)),
        strip.text = element_text(family = "JaneAusten", size = 12, margin = margin(b = 10)),
        axis.text.x = element_text(family = "Sylfaen", size = 10, hjust = 0.5),
        axis.text.y = element_markdown(family = "Sylfaen", size = 11, hjust = 1))
```  
\  

Some traits are pretty flippant, like Elizabeth Bennet's top trait, "treasure", which has "trash" at the other end of the spectrum. Keep in mind that this data is from an online quiz, not a rigorous scientific study -- it's just a bit of fun!

The traits for the rest of the characters are spot-on: **selfish** George Wickham, **kind** Jane, **foolish** Lydia, **awkward** Mr. Collins, and **rich** and **reserved** Mr. Darcy.  
\  


### *The Lion King*

```{r lion-king-example, fig.align = 'center', fig.width = 8, fig.height = 7.75}
lion_king_palette <- c("#FFBA08", "#FCA311", "#E85D04",
                       "#DC2F02", "#791901", "#38160D")

lion_king_example <- personalities %>%
  filter(!is_emoji,
         fictional_work == "The Lion King") %>%
  group_by(character_name) %>%
  top_n(8, abs(mean)) %>%
  ungroup() %>%
  mutate(spectrum_label = ifelse(mean > 0,
                                 glue("{spectrum_high}<br><span style='font-size:8pt'>({spectrum_low})</span>"),
                                 glue("{spectrum_low}<br><span style='font-size:8pt'>({spectrum_high})</span>")),
         mean = abs(mean),
         spectrum_label = reorder_within(spectrum_label, mean, character_name))

lion_king_example %>%
  ggplot() +
  geom_col(aes(mean, spectrum_label, fill = character_name), show.legend = FALSE) +
  geom_vline(xintercept = seq(0, 50, by = 10), colour = "#FFFBEB", size = 0.1) +
  scale_y_reordered() +
  scale_fill_manual(values = lion_king_palette) +
  expand_limits(x = c(0, 50)) +
  facet_wrap(~ character_name, scales = "free_y") +
  labs(title = "The Lion King",
       caption = "Visualization: @alexcookson | Data: Open Source Psychometrics Project") +
  theme(plot.title = element_text(family = "African", size = 30, hjust = 0.5, margin = margin(t = 10, b = 20)),
        plot.background = element_rect(fill = "#FFFBEB", colour = NA),
        plot.margin = margin(10, 10, 10, 10),
        plot.caption = element_text(family = "IBM Plex Sans", size = 8, colour = "grey50", margin = margin(t = 10)),
        strip.text = element_text(family = "African", size = 12, margin = margin(b = 10)),
        axis.text.x = element_text(family = "IBM Plex Sans", size = 10, hjust = 0.5),
        axis.text.y = element_markdown(family = "IBM Plex Sans", size = 11, hjust = 1))
```  
\  

Scar has great traits for a villain (including the very on-the-nose **villainous**!). Timon is **thin** and Pumbaa is **thick**. And Nala's top trait is **beautiful**, which I find a bit weird for an animated lion. (In fact, it gives some hints at how characters of different genders are portrayed and perceived in media, something I plan to analyze more in a future post.)  
\  


## Principal Component Analysis (PCA)

Now that we have a sense of the data, let's get into principal component analysis. There are some great resources out there already, like Julia Silge's [PCA walkthrough using hip hop songs](https://juliasilge.com/blog/best-hip-hop/) to see how it can be applied, [`tidymodels` documentation](https://recipes.tidymodels.org/reference/step_pca.html) to see how `tidymodels` approaches PCA, and StatQuest's [step-by-step guide to PCA](https://www.youtube.com/watch?v=FgakZw6K1QQ) to understand some of the underlying mathematics.  
\  


### Helper datasets

Before we get started, let's create create some helper datasets that will help us down the line. Why? We'll be conducting PCA with just code fields to identify characters and spectrums. Once the PCA is done, we'll want to inspect the results and it's not very informative to see character "HP/2" or spectrum "BAP209". Instead, we want to see "Harry Potter" or "city-slicker to country-bumpkin".

```{r helper-datasets}
character_list <- personalities %>%
  distinct(character_code, fictional_work, character_name)

spectrum_list <- personalities %>%
  distinct(spectrum, spectrum_low, spectrum_high)
```  
\  


### PCA with `tidymodels`

`tidymodels` makes PCA *very* straightforward -- in this case, it takes about 10 lines of code to set up the data properly and run the PCA.

First, let's convert our data to a format that's closer to a matrix than a dataframe. We want each row to correspond to a character and each column to correspond to a spectrum. We'll also have a special column with `character_code` so that we know who's who once we're done!

This is what the data we're feeding into `tidymodels` looks like:

```{r data-for-pca}
pca_data <- personalities %>%
  filter(!is_emoji) %>%
  select(character_code, spectrum, mean) %>%
  pivot_wider(names_from = spectrum, values_from = mean)

pca_data %>%
  head(6) %>%
  select(1:5) %>%
  kable(format = "html") %>%
  kable_styling()
```  
\  

Second, we'll create a recipe that specifies `character_code` as the `id` column, normalizes the scores (this is *very* important in PCA), and runs the PCA itself.


```{r pca}
# Write recipe for PCA
pca_recipe <- recipe(~ ., data = pca_data) %>%
  # Specify character_code as key/id column
  update_role(character_code, new_role = "id") %>%
  # Normalize sets mean to zero and standard deviation of one
  step_normalize(all_predictors()) %>%
  # PCA is done here
  # Use threshold to specify we want to capture 90% of variance in the data
  step_pca(all_predictors(),
           threshold = 0.9)

# prep() implements the recipe
# bake() applies the PCA model to a dataset
personality_pca <- prep(pca_recipe) %>%
  bake(new_data = pca_data)
```

That's it. The PCA is done. (Really!)  
\  


### What is a principal component?

You can think of each principal component (PC) as bundling many spectrums that tell us similar information into a single variable. Consider Mr. Darcy's top traits:

```{r mr-darcy-example, fig.align = 'center', fig.width = 7, fig.height = 5.75}
pride_example %>%
  filter(character_name == "Mr. Darcy") %>%
  mutate(spectrum_label = str_replace(spectrum_label, "8pt", "10pt"),
         spectrum_label = fct_reorder(spectrum_label, mean)) %>%
  ggplot() +
  geom_col(aes(mean, spectrum_label),
           fill = "#472411", show.legend = FALSE) +
  geom_vline(xintercept = seq(0, 50, by = 10), colour = "#FFF1E6", size = 0.1) +
  expand_limits(x = c(0, 50)) +
  scale_y_reordered() +
  labs(title = "Mr. Darcy") +
  theme(plot.title = element_text(family = "JaneAusten", face = "bold", size = 30, hjust = 0.5),
        plot.background = element_rect(fill = "#FFF1E6", colour = NA),
        plot.margin = margin(10, 10, 10, 10),
        axis.text.x = element_text(family = "Sylfaen", size = 12, hjust = 0.5),
        axis.text.y = element_markdown(family = "Sylfaen", size = 16, hjust = 1))
```  
\  

He's reserved, private, guarded, reclusive, confidential, and an introvert. All of these are getting at the same thing. If we already know that Mr. Darcy is **reserved**, do we learn that much more about him if we also know he's **private**? Not really, right? These two traits tend to go together (if you're curious, their correlation is 0.909).

Intuitively, *all* the reserved-type traits go together. So let's bundle them all together into a single, new variable. Why don't we call it, oh, I don't know, a ***principal component***? If we knew what Mr. Darcy's score was on this new principal component that bundles together reserved, private, etc., we'd still have a pretty good idea of what type of person he is, even though we wouldn't know his scores for each individual trait. *This is what PCA does*, except on a larger scale and with math instead of with intuition, as we just did with Mr. Darcy.  
\  


### What do the top principal components tell us?

We can see the composition of each principal component, which tells us which traits have been bundled together. Here are the top six PCs:

```{r top-components, fig.align = 'center', fig.width = 8, fig.height = 10}
# Custom colour palette
top_pc_palette <- c("#264653", "#2A9D8F", "#8AB17D",
                    "#E9C46A", "#F4A261", "#E76F51")

# Inspect a few principal components to see what they capture
prep(pca_recipe) %>%
  # Inspect the recipe after it has completed step 2
  tidy(type = "coef", number = 2) %>%
  # Add interpretable names of spectrums
  left_join(spectrum_list, by = c("terms" = "spectrum")) %>%
  # Take the end of the spectrum that the principal component value is closest to
  mutate(label = ifelse(value > 0,
                        glue("{spectrum_high} <span style='font-size:7pt'>({spectrum_low})</span>"),
                        glue("{spectrum_low} <span style='font-size:7pt'>({spectrum_high})</span>")),
         abs_value = abs(value)) %>%
  # Look at the first six components
  filter(component %in% paste0("PC", 1:6)) %>%
  # Look at the 8 strongest elements of each component
  group_by(component) %>%
  top_n(8, abs_value) %>%
  ungroup() %>%
  # Reorder spectrum labels so that they will be graphed from highest to lowest
  # reorder_within() is useful for when you want things ordered in a facetted graph
  mutate(text_label = label,
         label = reorder_within(label, by = abs_value, within = component)) %>%
  ggplot() +
  geom_col(aes(abs_value, label, fill = component),
           show.legend = FALSE) +
  geom_richtext(aes(x = 0, y = label, label = text_label),
            hjust = 0, nudge_x = 0.002, family = "IBM Plex Sans",
            colour = "white", size = 3.7, fontface = "bold",
            fill = NA, label.colour = NA) +
  scale_x_continuous(breaks = seq(0, 0.2, by = 0.05),
                     expand = c(0, 0)) +
  scale_y_reordered() +
  scale_fill_manual(values = top_pc_palette) +
  facet_wrap(~ component, scales = "free_y", ncol = 2) +
  labs(title = "What are the top principal components?",
       subtitle = paste("Each component is shown with the eight traits that contribute most to that component.",
                        "The opposing trait is shown in parentheses for context.",
                        sep = "\n")) +
  theme(text = element_text(family = "IBM Plex Sans"),
        plot.title = element_text(size = 24, margin = margin(t = 10)),
        plot.subtitle = element_text(size = 10, margin = margin(t = 10, b = 20)),
        plot.margin = margin(0, 20, 20, 20),
        panel.spacing.y = unit(20, "points"),
        strip.text = element_text(size = 12, hjust = 0, margin = margin(0, 5, 5, 5)),
        axis.text.x = element_text(size = 9, hjust = 0.5))
```  
\  

I won't rehash what's in the graph, so inspect it yourself. But I can imagine characters from some of my favourite books and shows fitting nicely into some of these PCs.

In fact, why don't we see which characters score highest and lowest among these components?

```{r pca-single-component-examples, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 10}
component_descriptions <- prep(pca_recipe) %>%
  tidy(type = "coef", number = 2) %>%
  filter(component %in% c(paste0("PC", 1:6))) %>%
  left_join(spectrum_list, by = c("terms" = "spectrum")) %>%
  group_by(component) %>%
  top_n(2, abs(value)) %>%
  mutate(spectrum_label = ifelse(value > 0, spectrum_high, spectrum_low),
         component = paste0("PC0", parse_number(component))) %>%
  summarise(component_description = paste(spectrum_label, collapse = ", "))

character_list %>%
  left_join(personality_pca, by = "character_code") %>%
  pivot_longer(starts_with("PC"), names_to = "component", values_to = "value") %>%
  group_by(component, sign = sign(value)) %>%
  top_n(4, abs(value)) %>%
  ungroup() %>%
  filter(component %in% c(paste0("PC0", 1:6))) %>%
  mutate(label = glue("{character_name}<br><span style='font-size:7pt'>({fictional_work})</span>"),
         label = reorder_within(label, by = value, within = component)) %>%
  left_join(component_descriptions, by = "component") %>%
  mutate(component_label = glue("{component}<br><span style='font-size:9pt'>(e.g., {component_description})</span>")) %>%
  ggplot(aes(value, label, fill = component)) +
  geom_col(show.legend = FALSE) +
  geom_vline(xintercept = 0, colour = "grey50") +
  facet_wrap(~ component_label, scales = "free_y", ncol = 2) +
  scale_y_reordered() +
  scale_fill_manual(values = top_pc_palette) +
  labs(title = "Who are the most extreme characters?",
       subtitle = paste("Each component shows its four highest- and lowest-scoring characters.",
                        "Below component titles are examples of the traits that comprise that component.",
                        sep = "\n")) +
  theme(text = element_text(family = "IBM Plex Sans"),
        plot.title = element_text(size = 24, margin = margin(t = 10)),
        plot.subtitle = element_text(size = 10, margin = margin(t = 10, b = 20)),
        plot.margin = margin(0, 20, 20, 20),
        panel.spacing.y = unit(20, "points"),
        strip.text = element_markdown(size = 14, hjust = 0.5, margin = margin(0, 5, 5, 5)),
        axis.text.x = element_text(size = 9, hjust = 0.5),
        axis.text.y = element_markdown(size = 9, hjust = 1))
```  
\  

These are spot-on! Joffrey Baratheon is ludicrous and deranged, Gus Fring is hard and armoured, and Leslie Knope is most definitely obsessed and hurried. Characters on opposite ends of the PCs are spot-on, too, from tiresome, stick-in-the-mud Mr. Collins to refined, manicured Tahani Al-Jamil.  
\  

We can also combine components to see how they relate to one another. (In fancier terminology, we're taking a bivariate view instead of a univariate view.) For example, if we plot `PC01` (deranged vs. resonable) against `PC04` (rugged vs. refined), we can see where characters fall along both components, resulting in a more nuanced look at their personalities. Characters farther to the right (or left) are more deranged (or reasonable). And characters higher up (or down) are more rugged (or refined).

```{r interactive-pca-scatterplot, fig.align = 'center', fig.width = 8, fig.height = 8}
p <- character_list %>%
  left_join(personality_pca, by = "character_code") %>%
  mutate(Character = character_name,
         `Fictional Work` = fictional_work,
         colour_group = case_when(PC01 > 0 & PC04 > 0 ~ "deranged_rugged",
                                  PC01 > 0 ~ "deranged_refined",
                                  PC04 < 0 ~ "reasonable_refined",
                                  TRUE ~ "reasonable_rugged"),
         alpha_factor = sqrt(PC01 ^ 2 + PC04 ^ 2)) %>%
  ggplot(aes(PC01, PC04, label = Character, label2 = `Fictional Work`)) +
  geom_hline(yintercept = 0, lty = 2, alpha = 0.8, colour = "grey50") +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.8, colour = "grey50") +
  geom_point(aes(colour = colour_group, alpha = alpha_factor)) +
  annotate("text", x = 14, y = 14, label = "Deranged\n+\nRugged",
           family = "IBM Plex Sans", size = 7, colour = "#cc0024") +
  annotate("text", x = 14, y = -14, label = "Deranged\n+\nRefined",
           family = "IBM Plex Sans", size = 7, colour = "#4b264d") +
  annotate("text", x = -14, y = -14, label = "Reasonable\n+\nRefined",
           family = "IBM Plex Sans", size = 7, colour = "#244F26") +
  annotate("text", x = -14, y = 14, label = "Reasonable\n+\nRugged",
           family = "IBM Plex Sans", size = 7, colour = "#016eae") +
  expand_limits(x = c(-21, 21),
                y = c(-16, 16)) +
  scale_colour_manual(values = c("#4b264d", "#cc0024", "#244F26", "#016eae")) +
  labs(title = "") +
  theme(legend.position = "none",
        text = element_text(family = "IBM Plex Sans"),
        plot.title = element_text(margin = margin(t = 10, b = 30)))

ggplotly(p, tooltip = c("label", "label2"))
```


Taking this view also emphasizes that **a single component can only take you so far**. There isn't just one type of "deranged" (`PC01`) character. Adding another component, in this case "rugged vs. refined" (`PC04`), we see that there are different flavours of deranged. For example, Jayne Cobb (*Firefly*) and Lucille Bluth (*Arrested Development*) have approximately the same score on `PC01`, but very different scores on `PC04`. I would agree that they're both deranged, but in very different ways. Adding a third (or fourth! or fifth!) component into the mix would give us even more nuance. (But I won't do that here, because visualizing more than two continuous dimensions gets tricky.)  
\  


### Conclusion

We've learned a lot about this dataset by conducting PCA and exploring the results. But dimensionality reduction techniques like PCA are often a precursor to other things, like developing classification models. I also think this dataset is so rich and interesting that you could explore it in other ways that have nothing to do with PCA.

If you'd like to take this analysis further, you can find the data on my [GitHub repo of interesting datasets](https://github.com/tacookson/data/tree/master/fictional-character-personalities). Some areas to look into are:

- How does gender affect how characters are portrayed and perceived?
- What are the clusters of personality types? Which characters from different fictional works are most like one another?
- Are certain personalities more prominent in certain genres? For example, do sitcoms tend to have personalities that other genres don't?
